---
title: "COVID-19 in Newfoundland and Labrador"
author: "John W"
date: "`r Sys.Date()`"
output: html_document
---

# Introduction

This notebook studies a public data set on COVID-19 that was published online
by Health Canada. The data set records a regular count of the total number of
cases and deaths attributed to the disease in each of the Canadian provinces. The
objective is to use this data set in order to study the impact of the disease
upon Newfoundland and Labrador.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,     # hide code
  # results = "hide", # hide printed output
  message = FALSE,  # hide messages
  warning = FALSE   # hide warnings
)
```

```{r}
  library(tidyverse)
  library(viridis)
  library(scales)

  df <- read.csv("https://health-infobase.canada.ca/src/data/covidLive/covid19-download.csv")
  # Convert all non-numeric values for the total case count to N/A
  df$totalcases <- as.numeric(df$totalcases)
  # Omit all records with N/A values
  df <- na.omit(df)
  df <- df %>%
    # Select province name, date and total case count
    dplyr::select(prname,date,totalcases,numdeaths) %>%
    # Filter out all provinces but Newfoundland and Labrador
    filter(prname == "Newfoundland and Labrador") %>%
    # Filter out any records with total case counts of zero
    filter(totalcases != 0)
  # Convert date values from strings into actual dates
  df$date <- as.Date(df$date)
```

# Analysis

## When were the first cases of COVID-19 officially reported in Newfoundland and Labrador?

```{r}
  df %>%
    slice(1)
```

The first six cases of COVID-19 in Newfoundland and Labrador were reported on 
**March 21, 2020.**

## What was the progression of COVID-19 in Newfoundland and Labrador?

```{r}
  df %>%
    ggplot(mapping=aes(x=date,y=totalcases)) +
    geom_line(linewidth=1,color=viridis(1)) +
    labs(x="Date", y="Total Case Count", title="COVID-19 in Newfoundland and Labrador", subtitle="Total Case Count")
```

The progression of COVID-19 was initially slow, with case numbers remaining in
the low thousands up until a point in 2022. This suggests that there was a fairly
limited spread of the disease during the early stages of the pandemic in the province.
A sharp period of growth in total case numbers began in 2022, likely indicating a
major wave of the disease at this time. From mid 2022 to 2023 there was a moderate
increase in COVID-19 cases, with the curve continuing to rise at a slower and
steadier pace. Case numbers plateau during the period between 2023 and 2024. The graph
demonstrates a classic pattern in pandemic growth; initial control, a sudden,
large outbreak, then finally a gradual stabilization as spread of the disease 
slowed over time.

## How many deaths occurred in Newfoundland and Labrador due to COVID-19?

```{r}
  df %>%
    slice_tail(n=1) %>%
    mutate(death_rate=numdeaths / totalcases) %>%
    mutate(death_rate=percent(death_rate, accuracy=0.01))
```

As of the last recording in May of 2024, the total number of deaths due to COVID-19
in the province was 412, only **0.70%** of the total case count.

## How does this number compare to the rest of the Canadian provinces?

```{r}
  canada_df <- read.csv("https://health-infobase.canada.ca/src/data/covidLive/covid19-download.csv")
  # Convert all non-numeric values for the total case count to N/A
  canada_df$totalcases <- as.numeric(canada_df$totalcases)
  # Omit all records with N/A values
  canada_df <- na.omit(canada_df)
  canada_df <- canada_df %>%
    # Select province name, date and total case count
    dplyr::select(prname,date,totalcases,numdeaths) %>%
    # Filter out any records with total case counts of zero
    filter(totalcases != 0)
  # Convert date values from strings into actual dates
  canada_df$date <- as.Date(canada_df$date)
  
  canada_df %>%
    arrange(prname, date) %>%
    group_by(prname) %>%
    slice_tail(n = 1) %>%
    ungroup() %>%
    mutate(death_rate = numdeaths / totalcases) %>%
    arrange(desc(death_rate)) %>%
    # Order the provinces by value instead of alphabetically
    ggplot(aes(x = reorder(prname, death_rate), y = death_rate, fill=death_rate)) +
    geom_col() +
    coord_flip() +
    scale_y_continuous(labels = percent) +
    scale_fill_viridis_c(labels=percent,name="Death Rate") +
    labs(title="Death Rates Due to COVID-19",subtitle="By Province",x="Province", y="Death Rate")
```

Across Canada, Newfoundland and Labrador had one of the lowest death rates due
to COVID-19 during the pandemic, at around **0.70%,** putting it well below
the top-rated province of British Columbia (with a death rate around **1.67%**).

# Conclusion

In general, then, it is apparent that in terms of total case numbers and deaths,
Newfoundland and Labrador experienced a relatively low impact due to the pandemic.
The province experienced one of the lowest death rates across Canada, indicating
relatively strong outcomes during the pandemic. This could be attributed to a
number of factors, such as healthcare response and prevention measures, as well as
population density and size. More study is needed to confirm the fundamental reasons
for Newfoundland's relatively strong performance during the pandemic.