---
title: "COVID-19 in Newfoundland and Labrador"
author: "John W"
date: "`r Sys.Date()`"
output: html_document
---

# Introduction

This notebook studies public datasets on COVID-19 that were published online
by Health Canada:

- [here](https://health-infobase.canada.ca/src/data/covidLive/covid19-download.csv)
- [here](https://health-infobase.canada.ca/src/data/covidLive/vaccination-coverage-byVaccineType.csv)

The objective is to use this data set in order to study the impact of the disease
upon Newfoundland and Labrador.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,     # hide code
  # results = "hide", # hide printed output
  message = FALSE,  # hide messages
  warning = FALSE   # hide warnings
)
```

```{r}
  library(tidyverse)
  library(viridis)
  library(scales)

  df <- read.csv("https://health-infobase.canada.ca/src/data/covidLive/covid19-download.csv")
  df$totalcases <- as.numeric(df$totalcases)
  df <- na.omit(df)
  df <- df %>%
    dplyr::select(prname,date,totalcases,numdeaths) %>%
    filter(prname == "Newfoundland and Labrador") %>%
    filter(totalcases != 0)
  df$date <- as.Date(df$date)
```

# Analysis

## When were the first cases of COVID-19 officially reported in Newfoundland and Labrador?

```{r}
  df %>%
    slice(1)
```

The first six cases of COVID-19 in Newfoundland and Labrador were reported on 
**March 21, 2020.**

## What was the progression of COVID-19 in Newfoundland and Labrador?

```{r}
  df %>%
    ggplot(mapping=aes(x=date,y=totalcases)) +
    geom_line(linewidth=1,color=viridis(1)) +
    labs(x="Date", y="Total Case Count", title="COVID-19 in Newfoundland and Labrador", subtitle="Total Case Count")
```

The progression of COVID-19 was initially slow, with case numbers remaining in
the low thousands up until a point in 2022. This suggests that there was a fairly
limited spread of the disease during the early stages of the pandemic in the province.
A sharp period of growth in total case numbers began in 2022, likely indicating a
major wave of the disease at this time. From mid 2022 to 2023 there was a moderate
increase in COVID-19 cases, with the curve continuing to rise at a slower and
steadier pace. Case numbers plateau during the period between 2023 and 2024. The graph
demonstrates a classic pattern in pandemic growth; initial control, a sudden,
large outbreak, then finally a gradual stabilization as spread of the disease 
slowed over time.

## How many deaths occurred in Newfoundland and Labrador due to COVID-19?

```{r}
  df %>%
    slice_tail(n=1) %>%
    mutate(death_rate=numdeaths / totalcases) %>%
    mutate(death_rate=percent(death_rate, accuracy=0.01))
```

As of the last recording in May of 2024, the total number of deaths due to COVID-19
in the province was 412, only **0.70%** of the total case count.

## How does this number compare to the rest of the Canadian provinces?

```{r}
  canada_df <- read.csv("https://health-infobase.canada.ca/src/data/covidLive/covid19-download.csv")
  canada_df$totalcases <- as.numeric(canada_df$totalcases)
  canada_df <- na.omit(canada_df)
  canada_df <- canada_df %>%
    dplyr::select(prname,date,totalcases,numdeaths) %>%
    filter(totalcases != 0)
  canada_df$date <- as.Date(canada_df$date)
  
  canada_df %>%
    arrange(prname, date) %>%
    group_by(prname) %>%
    slice_tail(n = 1) %>%
    ungroup() %>%
    mutate(death_rate = numdeaths / totalcases) %>%
    arrange(desc(death_rate)) %>%
    ggplot(aes(x = reorder(prname, death_rate), y = death_rate, fill=death_rate)) +
    geom_col() +
    coord_flip() +
    scale_y_continuous(labels = percent) +
    scale_fill_viridis_c(labels=percent,name="Death Rate") +
    labs(title="Death Rates Due to COVID-19",subtitle="By Province",x="Province", y="Death Rate")
```

Across Canada, Newfoundland and Labrador had one of the lowest death rates due
to COVID-19 during the pandemic, at around **0.70%,** putting it well below
the top-rated province of British Columbia (with a death rate around **1.67%**).

## What is the total number of fully vaccinated individuals by vaccine primary series in Newfoundland and Labrador?

```{r}

data <- read.csv("https://health-infobase.canada.ca/src/data/covidLive/vaccination-coverage-byVaccineType.csv")

plot_data <- data |>
  clean_names() |>
  filter(prename == "Newfoundland and Labrador") |>
  group_by(primary_series) |>
  summarise(total_fully = max(numtotal_fully, na.rm = TRUE))

ggplot(plot_data, aes(x = reorder(primary_series, total_fully), y = total_fully, fill = total_fully)) +
  geom_col() +
  theme_minimal() +
  labs(
    x = "Primary Series",
    y = "Total Fully Vaccinated",
    title = "Fully Vaccinated Counts by Vaccine Primary Series",
    subtitle = "Newfoundland and Labrador (2020-2024)",
    fill = "Total Fully Vaccinated"
  ) +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 20)) +
  scale_y_continuous(labels = scales::comma) +
  theme(title = element_text(hjust = 0.5)) +
  coord_flip() +
  scale_fill_viridis_c()
```

This horizontal bar chart compares the total number of fully vaccinated individuals in Newfoundland and Labrador by vaccine primary series. The Pfizer-BioNTech (Comirnaty) series clearly has the highest total, standing well above all other categories and indicating it was the most widely used vaccine. Mixed vaccine series (both mRNA mixed and unspecified mixed) form the next largest groups, suggesting that a notable portion of the population received combinations of vaccines. Moderna (Spikevax) also shows substantial uptake, though it trails behind Pfizer and the mixed categories. AstraZeneca and AstraZeneca–mRNA combinations appear at much lower levels, indicating more limited use. Other categories, such as Janssen (Johnson & Johnson), Novavax, and unknown or other mixed series, have very small totals, reflecting minimal contribution to overall vaccination.

Overall, the plot shows that mRNA vaccines—particularly Pfizer—dominated the vaccination rollout, with mixed-dose strategies also playing a significant role, while non-mRNA vaccines were used far less frequently.

# Conclusion

The data from Newfoundland and Labrador shows a clear and typical pandemic trajectory, beginning with a slow initial spread after the first reported cases in March 2020, followed by a major surge in 2022 and then a gradual stabilization of case numbers into 2023–2024. Despite this increase in total cases over time, the province maintained a relatively low COVID-19 death rate of about 0.70%, which is among the lowest in Canada and suggests effective public health response and healthcare capacity throughout the pandemic.

The vaccination data further helps explain this outcome. The rollout was heavily dominated by mRNA vaccines, particularly Pfizer-BioNTech, with Moderna also contributing substantially, while AstraZeneca, Janssen, and Novavax made up much smaller proportions of total vaccinations. The presence of mixed vaccine series indicates flexibility in vaccine administration strategies and widespread population coverage. Taken together, the vaccination patterns and relatively low mortality rate suggest that high uptake of mRNA vaccines, combined with public health measures, played an important role in limiting severe outcomes even as case numbers eventually increased.